---
title: "Data reading and processing"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 5
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Data reading and processing}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---


```{r message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls())

library(REDCapDM)
library(dplyr)
library(purrr)
library(knitr)
library(kableExtra)
```

<br>

This vignette provides a summary of the straightforward and common use of [REDCapDM](https://github.com/bruigtp/REDCapDM) to interact with [REDCap](https://www.project-redcap.org/) data.

<br>

# **Read REDCap data into R**

`REDCapDM` allows you to import REDCap data either from local files or via an API connection. Both methods return a structured list containing your dataset, dictionary, and event-form mapping (for longitudinal projects).

## **1. Using local files**

Ensure you have exported the following files from REDCap and placed them in the same directory:

1. **R data file** – contains your REDCap dataset (`.rds` or `.RData`)

2. **Dictionary file** – CSV file containing variable metadata

3. **Event-form mapping file** (only for longitudinal projects) – CSV describing which forms belong to each event

Then, provide the paths to these files using the `redcap_data()` function:

```{r message=FALSE, warning=FALSE, comment=NA, eval=FALSE}
dataset <- redcap_data(
  data_path = "C:/Users/username/example.r",
  dic_path = "C:/Users/username/example_dictionary.csv",
  event_path = "C:/Users/username/events.csv" # optional for longitudinal projects
)
```

⚠️ Tip: If your project is not longitudinal, you can omit `event_path`.


## **2. Using API connection**

You can also import data directly from REDCap using your project’s **API token** and **URL**:

```{r eval=FALSE, message=FALSE, warning=FALSE, comment=NA}
dataset_api <- redcap_data(
  uri = "https://redcap.com/api/",
  token = "YOUR_API_TOKEN_HERE"
)
```

This method automatically retrieves the event-form mapping for longitudinal projects.

⚠️ Tip: Keep your API token secure and do not share it publicly.

## **3. Understanding the output**

[`redcap_data()`](https://bruigtp.github.io/REDCapDM/reference/redcap_data.html) returns a list with three elements:

| Element      | Description                                                              |
| ------------ | ------------------------------------------------------------------------ |
| `data`       | Imported dataset containing all REDCap variables                         |
| `dictionary` | Metadata for each variable, including labels, types, and branching logic |
| `event_form` | Mapping between events and forms (only for longitudinal projects)        |


<br>
<br>

# **Process and transform REDCap data**

After importing your dataset into R, `REDCapDM` provides several functions to **process and transform your data** efficiently. The main function is [`rd_transform()`](https://bruigtp.github.io/REDCapDM/reference/rd_transform.html), which performs multiple operations in one step.

## **1. Core transformations**

`rd_transform()` performs the following actions:

| Transformation | Description |
|----------------|-------------|
| Variable deletion | Remove specific variables or patterns |
| Recalculate fields | Update REDCap calculated fields |
| Checkbox transformation | Convert checkbox variables into separate variables with option names |
| Factor conversion | Replace original variables with their factor version |
| Branching logic conversion | Translate REDCap logic into R logic for easier processing |


## **2. Basic usage**

The minimum required inputs are the dataset and dictionary. For longitudinal projects, include the event-form mapping for full functionality.

You can provide these inputs either as a **single list** (output of `redcap_data()`) or as **separate arguments**:

```{r message=FALSE, warning=FALSE, comment=NA}
#Option A: Provide the full project list
covican_transformed <- rd_transform(covican)

#Option B: Provide individual arguments
covican_transformed <- rd_transform(
  data = covican$data, 
  dic = covican$dictionary, 
  event_form = covican$event_form
)
```

The output is a list containing:

- **data**: Transformed dataset

- **dictionary**: Updated variable dictionary

- **event_form**: Event-form mapping (if applicable)

- **results**: Summary of all transformations

You can view the transformation results:

```{r message=FALSE, warning=FALSE, comment=NA}
#Print the results of the transformation
covican_transformed$results
```

## **3. Splitting by event**

For longitudinal projects, you can split the transformed dataset by event using the `final_format = "by_event"` argument:

```{r message=FALSE, warning=FALSE, comment=NA}
dataset <- rd_transform(covican,
                        final_format = "by_event")
```

Where the transformed dataset is a tibble object, containing data frames for each event in the REDCap project.

```{r message=FALSE, warning=FALSE, comment="#>", collapse = TRUE}
dataset$data
```

## **4. Splitting by form**

Or, alternatively, it can be split by form using `final_format = "by_form"` argument:

```{r message=FALSE, warning=FALSE, comment=NA}
dataset <- rd_transform(covican,
                        final_format = "by_form")
```

Where the tibble object is composed by data frames corresponding to each form in the REDCap project.

```{r message=FALSE, warning=FALSE, comment="#>", collapse = TRUE}
dataset$data
```

<br>
<br>

# **Quick troubleshooting checklist**

1. Did you export the variable dictionary from REDCap? Without it, label
   conversions and checkbox expansions are error-prone.
2. Are API calls failing? Verify `uri`, `token` and your project's export
   permissions.

**For more information, consult the complete vignette available at: https://bruigtp.github.io/REDCapDM/articles/REDCapDM.html**

<br>
<br>
